Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2024-04-17 response_r2.md

---
title: Response to Reviewers (Second Round)
subtitle: "COATi: statistical pairwise alignment of protein-coding sequences"
author:
	- Juan J. García Mesa
	- Ziqi Zhu
	- Reed A. Cartwright
mainfont: TeX Gyre Pagella
geometry:
	- margin=1in
---

<!-- pandoc --pdf-engine=lualatex -o response_r1.pdf response_r1.md -->

Dear Editor,

We are resubmitting our manuscript, "COATi: statistical pairwise alignment of protein coding sequences", for your consideration. This manuscript describes a new model for aligning protein coding sequences which produces more accurate alignments, while being robust to common artifacts found in genomic annotations. Our approach is implemented in the software package, COATi, which is released via an MIT license and available on GitHub. While COATi can generate multiple-sequence alignments, this paper is focused on our pairwise aligner which is has received more attention and validation. Additionally, to promote open science and reproducibility all data, procedures, and scripts used to validate COATi and compare it with alternative methods have been uploaded to GitHub.

We would like to thank the associate editor and two reviewers for their helpful comments and suggestions. We have addressed all of the concerns brought to our attention by the reviewers as described below.

Sincerely,  
Reed A. Cartwright, on behalf of all authors

## Reviewer 1 Comments

1. Comment: *It is now very clear and easy to read manuscript and a important edition to the alignment literature since it does well relative to the objective it pursues.  It has moved away from statistical alignment formulation in the first submission toward doing well in agreeing with benchmark alignments.  Some discussion of this would be useful.*

    TODO

1. Comment: *And how will the authors turn this into a multiple alignment tool?*

    COATi currently has an early-stage module for estimating pairwise alignments. However, it needs more work (and an additional round of funding) before it meets the high-quality standards exhibited by the pairwise aligner. We have included additional details about COATi MSA in the discussion.

## Review 2 Comments

1. Comment: *This is my second review of the manuscript “COATi: statistical pairwise alignment of protein-coding sequences” by García Mesa et al. and I have to say that it is practically a new manuscript. The first version really excited me, and I felt the science was both sound and very useful to the community, but the paper was difficult to understand. This version is much better; it is well-written, and it confirmed my belief that COATi is an important and useful contribution to the field of sequence alignment.*

    We appreciate your positive words.

1. Comment: *My only somewhat more conceptual comment is about the use of K2P distances. I buy the argument that “… the Kimura’s 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences” (p. 8, line 191) I am a bit mixed about its use. To be honest, I don’t think there are any truly appropriate distances for coding regions. However, p-distances would actually be more intuitive. The authors should consider calculating p-distances and placing them in the supporting information (or even replacing the K2P data). I don’t think this will change conclusions, but it would make things more intuitive for readers. Since I am certain this won’t change conclusions, I don’t want to push it that hard.*

    We calculated p-distances and have included the results in the supplement. The results were equivalent to K2P distances, which we now mention in the manuscript.

2. Comment: *I am not certain why the authors use “phase-3” indels rather than “phase-0”. I find phase-0 more intuitive, and it strikes me that this could be changed easily and would simplify the writing – the authors wouldn’t have to explain that they using phase-3 instead of phase-0!*

    We originally used phase-0 in our terminology. However, in our experience from presenting this research to diverse audiences, phase-3 was easier terminology to explain than phase-0. For example, "a phase-X gap occurs after the X position in a codon" is easier to explain than "a phase-X gap occurs before the X+1 position in a codon". Importantly, by using phase-3, we can succinctly explain the relationship between the frequencies of the gaps as phase-3 > phase-2 > phase-1. We have added additional language to the manuscript to emphasize these points.

3. Comment: *The authors should use “see” rather than c.f. on p. 5, line 82. The abbreviation should be cf., not c.f., and it should be used to mean “compare.” Given the Ranwez et al. references I don’t think the authors of this paper mean “compare.”*

    We do mean "compare" here because COATi's innovation is in similar spirit to MACSE but fundamentally different.

4. Comment: *I would consider adding the fact that their “…COATi results estimate that only 41% indels in protein-coding sequences between humans and gorillas are phase-3 indels” (on p. 15, line 370) to the abstract. This is a nice empirical demonstration that their methods is producing “biologically-reasonable” alignments.*

    Thank you for the suggestion. We have reworked the abstract and included this result.

5. Comment: *I’m a bit confused by the authors’ use of “frameshifts” when discussing the behavior of Clustal Omega (p. 15, line 402). A simple phase-1 or phase-2 indel that is a multiple of 3 is not a frameshift. Are they referring to separate indels that might have an intermediate that has a genuine frameshift? By “genuine frameshift” I mean a case where a presumably inactivating indel mutation that is not a multiple of 3 occurs and persists in a population long enough to have a nearby compensatory mutation that puts the sequence back in frame (i.e., the sum of then indel lengths is a multiple of 3). I think a minor rewrite for clarity would help readers.*

    In this instance, we are referring to an abiological frameshift that produces an abiological amino-acid translation. We have reworked this section for clarify our meaning.

6. Comment: *Overall, I really like this new version of the manuscript!*

    Thank you.

### 2024-04-17 alignpair_letter.tex
Update manuscript

\begin{abstract}
\noindent
Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation. Sequence artifacts and errors made during alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies. For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences that have been annotated in reference genomes. While such errors are eventually fixed in the reference genomes of model organisms, many genomes used by researchers contain these artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results. To address this need, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data. COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments.
\end{abstract}

@@

Within protein-coding sequences, indels may occur in between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}). Gaps that occur after the first position or second position of a codon are known as phase-1 and phase-2 gaps, respectively. Gaps that occur between codons are known as either phase-3 gaps (this study) or phase-0 gaps \citep[e.g.][]{taylor2004occurrence}. While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps. Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence (Fig.\ \ref{fig:aln}).

@@

COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; c.f.][]{holmes2001evolutionary}. We call this transducer the COATi FST. Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics. COATi implements both the \cite{muse_gaut_1994} codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
It also lets the user provide a codon substitution matrix. A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons \citep[c.f.][]{ranwez_macse_2011,ranwez_macse_2018}. This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.

@@

In order to use the COATi FST to align an output sequence against an input sequence, we first convert each sequence into an acceptor, represented as a linear transducer where the input and output symbols of each transition are identical and each transition represents one nucleotide of a sequence \citep{allauzen2007openfst}. By composing the input and output acceptors with the COATi FST, we generate a transducer of all possible alignments of the two sequences. Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment. If more that one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm. All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}. An example of an FST-based alignment can be found in Supplementary Materials Figure S1.

@@

Alignment inference impacts the estimation of evolutionary distances. To quantify the impact of aligners on the estimation of evolutionary distances, we calculated Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} for each sequence-pair and aligner combination. K2P corrects for multiple mutations at a site and takes into account differences in transition and transversion rates. It also assumes equal nucleotide frequencies and no variation in the rate of substitution across sites. While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences. We used the R software package ape \citep{paradis2019ape} to calculate K2P distances.
%
Since aligners influence evolutionary distances based on their tendencies to insert gaps into sequences, we also quantified the phases and lengths of gaps introduced by each method as well as the fraction of nucleotides that are aligned against a match, mismatch, or gap.

@@

% The simulation algorithm can introduce a pairwise alignment pattern to any two nucleotide sequences of equal length. The alignment pattern is given as a CIGAR string (Compact Idiosyncratic Gapped Alignment Report), a format commonly used to summarize aligned reads to a reference genome. Assigning one of the sequences as the reference, to distinguish between insertions and deletions, CIGAR strings can also summarize pairwise alignments by grouping the number of contiguous matches or mismatches `M', deletions `D', and insertions `I'. The resulting pattern combines these letters preceded by the number of characters for each section as they appear in the alignment. This pattern is introduced by replacing nucleotides with gaps as indicated by deletions on one sequence and randomly introducing residues where the CIGAR strings indicated insertions.

% Several safety checks are in place to ensure the algorithm runs correctly and the result is accurate. The assertions are divided into checking lengths and maintaining the reading frame of each section. The simulation can fail if the length of the sequences is different or if, without counting insertions, the length of the pattern to be inserted is longer. In addition, maintaining the phases of each section in the CIGAR string is important to avoid introducing errors such as frameshifts.

% I created the benchmark of alignments by using an equal number of randomly sampled gap patterns from each aligner.
% % After creating the dataset, I removed the gaps and measured how well different aligners were able to retrieve the true alignments.
% I used the dataset to evaluate the accuracy of COATi and a suite of popular aligners spanning various alignment methods:
% Clustal$\Omega$ v1.2.4 \citep{clustal_omega_sievers_2011},
% MACSE v2.06 \citep{ranwez_macse_2011}, MAFFT v7.505
% \citep{katoh2013mafft}, and PRANK v.150803 \citep{prank_loytynoja_2014}.

@@

% The computation of $d_{seq}$ involves characterizing the gaps present in the alignment. Then, a site-wise homology set $H(A)^i_j$ is calculated for each alignment $A$, sequence $i$, and character $j$. The distance between two alignments $A$ and $B$ is the average across all characters of the symmetric difference (or Hamming distance), represented as `$\triangle$' between homology sets over the length of such sets:

% \begin{equation}
% d(A,B) = \frac{1}{c} \sum_i \sum_j \frac{|H(A)^i_j \triangle H(B)^i_j|}{|H(A)^i_j|+|H(B)^i_j|}
% \end{equation}

% \noindent where $c$ is the sum of the sequence lengths.

@@

% \cyan{The first sentence intro is a good idea, but rewrite!}
% Sequencing artifacts and errors made in alignment can lead to inaccurate results in genomics. This is especially true in studies of positive selection and positively selected genes.

% Ks and Ka are, respectively, the number of substitutions per synonymous site and per non-synonymous site between two protein-coding genes.

% First, this method takes two aligned homologous protein-coding sequences and classifies the nucleotide sites in a sequence as nondegenerate, twofold degenerate, and fourfold degenerate. A site is nondegenerate if all possible changes at that site are nonsynonymous, twofold degenerate if one of the three possible changes is synonymous, and fourfold degenerate if all possible changes are synonymous. Second, the nucleotide changes between the two sequences are counted and divided as transitional (A$\leftrightarrow$G, C$\leftrightarrow$T) and transversional (\{A, G\}$\leftrightarrow$\{C, T\}). Third, the Kimura two-parameter distance \citep{kimura1980simple} is used to estimate the number of transitions and transversions per site type (nondegenerate, twofold degenerate, and fourfold degenerate), which is used as a correction factor for multiple hits. Finally, $k_s$ is the estimate of the average transitional rate at twofold and fourfold degenerate sites, and $k_a$ is the estimate of the average transversional rate at nondegenerate and twofold sites. In the results, these metrics are reported as the F$_1$ score, which is the harmonic mean of precision (true positives over total positives) and recall (true positives over true positives and false negatives). This score ranges between 0 and 1, with a score of 1 representing a perfect result.

% e compared the estimated evolutionary distance between the reference and inferred alignments using the Kimura's 2-parameter distance \citep{kimura1980simple}. The calculation of the Kimura 2-parameter model involves considering the rates of nucleotide transitions and transversions. The formula used for distance calculation is $D = -0.5 \cdot \log((1 - 2P - Q) \cdot \sqrt{1 - 2Q})$, where $P$ represents the proportion of transitional substitutions and $Q$ represents the proportion of transversional substitutions.

@@

The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable for human-gorilla gene pairs, with means of 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a). Conversely, the means for ClustalΩ, MACSE, and MAFFT were much higher, 1.5\%--1.8\%. These larger means are due to the fact that the right tails of the distributions for ClustalΩ, MACSE, and MAFFT were larger than the other methods. As expected, the alignment methods that ``overaligned'' sequences (produced more mismatches and fewer gaps) also produced higher evolutionary distances on average (Tab.~\ref{tab:gaps-empirical} and Fig.~\ref{fig:k2p-empirical}a). The differences between the distances inferred by COATi and other methods also shows ClustalΩ, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
%
COATi's alignments produced significantly lower K2P distances than ClustalΩ, MACSE, MAFFT, and PRANK. Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$. Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test). Additional analysis of the empirical alignments can be found in Supplementary Methods.

@@

COATi produced better alignments than ClustalΩ, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set. It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
%
COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}). All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests. (P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$). Notably, the average alignment error of the second best protocol was four times larger than COATi's. We also calculated the average distance of alignments between aligners and used principle coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}). COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions. Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for each source of gap patterns (Supplementary Materials, Fig.~S5). While all aligners, except ClustalΩ, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.

@@

\section*{Discussion}
% Using 16000 human genes and their gorilla orthologs from the ENSEMBL
% database \citembe{ensembl_hubbard_2002}, we simulated a data set of pairwise
% alignments with empirical gap patterns.
% We used the data set to evaluate the accuracy of COATi and a suite of popular aligners spanning multiple alignment methods:
% Clustal$\Omega$ v1.2.4 \citembe{clustal_omega_sievers_2011},
% MACSE v2.06 \citembe{ranwez_macse_2011}, MAFFT v7.505
% \citembe{katoh2013mafft}, and PRANK v.150803 \citembe{prank_loytynoja_2014}.

% After downloading, we removed 2232 sequence-pairs longer than 6000 nucleotides and then aligned the remaining pairs with all five methods.
% At least one aligner added gaps to 6048 sequence pairs, and no aligner added gaps to 7719 sequence pairs.
% We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
% The accuracy of inferred alignments compared to the benchmarks was measured using \citeauthor{metrics_blackburne_whelan_2011}'s \citeyearpar{metrics_blackburne_whelan_2011} distance metric, $d_{seq}$.
% Additionally, because different alignments may be evolutionary equivalent, we scored alignments using the codon-marginal-mg model, and any alignment that had the same score as the benchmark alignment was considered equivalent to the benchmark alignment.
% The accuracy of identifying positive and negative selection was calculated using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
% \citembe{ka_ks_li_1993} (Supplementary Methods).
% The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 1 representing a perfect result.

% NOTES:
% - Benchmark only contains seqeunces with gaps, which will make alignes look more different than they are in practice
% - simulalation contained a lot of frameshifts that produed stop codons. Probably too many.
% - redo colors of figures 1 - 3.
% - define an equivalent alignment again in the benchmark results

@@

ClustalΩ's amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts; however, it was the worst performing protocol that we tested. It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.  --- It even performed poorly on its own gap patterns. --- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our data set and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.

@@

We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols. Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S6). Different aligners tended to spread out from one another and in different directions from the benchmark. This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''. While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK also did well on ClustalΩ patterns. COATi did well on all patterns.

% COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences. To test how well COATi performs when the roles of sequences are reversed, we realigned the 7761 benchmark sequence alignments using gorilla as the reference. Notably, COATi was only able to align 4003 sequence pairs due to the presence of early stop codons or other artifacts in 3758 simulated gorilla sequences. While the simulation algorithm prevents disrupting the reading frame and introducing frameshifts, it does not prevent early stop codons from being formed in the descendant sequence. Despite this limitation, we analyzed the 4003 alignments and compared the results across methods, including COATi using the human sequence as the reference. The results show a decrease, albeit small, in accuracy across all metrics when the low-quality sequence is used as the reference in comparison to the reverse (Supplementary Materials Tab. 6). However, the results for COATi continue to be a significant improvement over other aligners.

COATi uses finite-state transducers to align pairs of sequences using a codon-aware statistical model. While COATi has multiple modes and models, here we have focused on evaluating the utility of COATi FST for estimating pairwise alignments. We have shown that COATi offers a biologically significant improvement over other methods when aligning pairs of sequences separated by short evolutionary distances in the presence of genomic artifacts. In other work, we have begun to explore application of COATi to more divergent sequences and the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.

COATi is under active development. We plan on extending it to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighing gap openings to reflect known selection on indel phases \citep{zhu2022profiling}. We plan on improving its multiple-sequence alignment and alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes. Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.

\section*{Availability}
The source code for COATi, along with documentation, is freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is implemented in C++. COATi is released under an MIT license. Additional information, code, and workflows to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing} and \url{https://github.com/jgarciamesa/alignpair_letter}.

>>>>

\begin{abstract}
\noindent
Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation. Sequence artifacts and errors made during alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies. For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences that have been annotated in reference genomes. More significantly, while indels can occur both within and between codons in natural sequences, most amino-acid- and codon-based aligners assume that indels only occur between codons. This mismatch between biology and alignment algorithms produces sub-optimal alignments and errors in downstream analyses.

To address these issues, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data. COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments. COATi can infer indels both within and between codons, leading to improved sequences alignments. We applied COATi to a dataset containing orthologous protein-coding sequences from humans and gorillas, and conclude that 41\% of indels occurred between codons, agreeing with an earlier study in rats and mice. We also applied to semi-empirical benchmark alignments and find that COATi outperforms several popular alignment programs via several measures.

\end{abstract}

%While such errors are eventually fixed in the reference genomes of model organisms, many genomes used by researchers contain these artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.

@@

Within protein-coding sequences, indels may occur in between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}). Gaps that occur after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively. Notably, phase-3 gaps occur between codons and are also known as phase-0 gaps \citep[e.g.][]{taylor2004occurrence}. Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2 which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}. While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps. Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence (Fig.\ \ref{fig:aln}).

@@

COATi aligns pairs of sequences using a statistical alignment model, which is implemented as a finite-state transducer derived from the composition of multiple FSTs, each representing a specific biological or technical process: (1) the codon substitution FST, (2) the indel FST, (3) the error FST, and (4) the ambiguity FST \citep[Figs.~\ref{fig:base-calling}--\ref{fig:coati-fst}; cf.][]{holmes2001evolutionary}. We call this transducer the COATi FST. Codon substitution models are uncommon in sequence aligners, despite their extensive use in phylogenetics. COATi implements both the \cite{muse_gaut_1994} codon model (codon-triplet-mg) and the Empirical Codon Model \citep[codon-triplet-ecm;][]{kosiol_ECM_2007}.
It also lets the user provide a codon substitution matrix. A key innovation of COATi is that it combines a codon substitution model with a nucleotide-based indel model, allowing gaps to occur both between and within codons \citep[cf.][]{ranwez_macse_2011,ranwez_macse_2018}. This also allows the aligner to be robust against sequencing artifacts that produce sequences with disrupted reading frames.

@@

In order to use the COATi FST to align an output sequence against an input sequence, we first convert each sequence into an acceptor, represented as a linear transducer where the input and output symbols of each transition are identical and each transition represents one nucleotide of a sequence \citep{allauzen2007openfst}. By composing the input and output acceptors with the COATi FST, we generate a transducer of all possible alignments of the two sequences. Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment. If more that one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm. All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}. An example of an FST-based alignment can be found in Supplementary Materials, Figure S1.

@@

Alignment inference impacts the estimation of evolutionary distances. To quantify the impact of aligners on the estimation of evolutionary distances, we calculated Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} for each sequence-pair and aligner combination.  K2P distances correct for multiple mutations at a site and takes into account differences in transition and transversion rates. K2P also assumes equal nucleotide frequencies and no variation in the rate of substitution across sites. While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences. We also calculated p-distances \citep{saitou1987neighbor} for each sequence-pair and aligner combination, which are simply the proportions of sites that differ between two sequences. We used the R software package ape \citep{paradis2019ape} to calculate the distances.
%
Since aligners influence evolutionary distances based on their tendencies to insert gaps into sequences, we also quantified the phases and lengths of gaps introduced by each method as well as the fraction of nucleotides that are aligned against a match, mismatch, or gap.

@@

@@

@@

@@

The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable for human-gorilla gene pairs, with means of 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a). Conversely, the means for ClustalΩ, MACSE, and MAFFT were much higher, 1.5\%--1.8\%. These larger means are due to the fact that the right tails of the distributions for ClustalΩ, MACSE, and MAFFT were larger than the other methods. As expected, the alignment methods that ``overaligned'' sequences (produced more mismatches and fewer gaps) also produced higher evolutionary distances on average (Tab.~\ref{tab:gaps-empirical} and Fig.~\ref{fig:k2p-empirical}a). The differences between the distances inferred by COATi and other methods also shows ClustalΩ, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
%
COATi's alignments produced significantly lower K2P distances than ClustalΩ, MACSE, MAFFT, and PRANK. Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$. Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test). Additional analysis of the empirical alignments can be found in Supplementary Methods, including results using p-distances, which did not differ from the results using K2P distances.

@@

COATi produced better alignments than ClustalΩ, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set. It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
%
COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}). All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests. (P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$). Notably, the average alignment error of the second best protocol was four times larger than COATi's. We also calculated the average distance of alignments between aligners and used principle coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}). COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions. Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for each source of gap patterns (Supplementary Materials, Fig.~S6). While all aligners, except ClustalΩ, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.

@@

\section*{Discussion}

@@

ClustalΩ's amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts. Because this approach depends on amino-acid translations, if a coding sequence contains an abiological frameshift, then translation will produce an abiological amino-acid sequence resulting in a poor alignment. ClustalΩ was the worst performing protocol that we tested. It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.  --- It even performed poorly on its own gap patterns. --- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our data set and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.

@@

We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols. Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7). Different aligners tended to spread out from one another and in different directions from the benchmark. This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''. While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK also did well on ClustalΩ patterns. COATi did well on all patterns.

COATi uses finite-state transducers to align pairs of sequences using a codon-aware statistical model, and the present study has shown that COATi offers a biologically significant improvement over other methods when aligning pairs of sequences separated by short evolutionary distances and in the presence of genomic artifacts. While COATi has multiple models, we have focused on evaluating the utility of COATi's complex finite-state-transducer for estimating pairwise alignments. In other work, we have begun to explore application of COATi to more divergent sequences and the accuracy of the marginal approximation of the COATi FST \citep{garcia2023dissertation}.

This study has focused on COATi's module for estimating pairwise alignments; however, COATi has additional tools, including modules to sample pairwise alignments from alignment space and generate multiple-sequence alignments. We have used the sampling module in a large study \citep{zhu2022profiling} and believe that it is ready to be used in further studies. On the other hand, COATi's multiple-sequence alignment algorithm is still in early development and needs additional work to ensure that it produces high quality alignments. Currently, COATi generates multiple sequence alignments by aligning non-reference sequences against a reference sequence and collapsing insertions following a guide tree.

COATi is under active development, and future work will enhance the multiple-sequence aligner to refine the initial alignment according to a scoring function consistent with COATi's FST model. We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighing gap openings to reflect known selection on indel phases \citep{zhu2022profiling}. We plan on improving its alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes. Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.

\section*{Availability}

The source code for COATi, along with documentation, is freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is implemented in C++. COATi is released under an MIT license. Additional information, code, and workflows to replicate the analysis can be found on GitHub: \url{https://github.com/jgarciamesa/coati-testing} and \url{https://github.com/jgarciamesa/alignpair_letter}.
